Abundance/diversity GAM

### 2020-09-21 analyses/abund_div_GAM.Rmd
update prior to trying to recover from a broken projecy

### FL and BB abundance / diversity
```{r echo=FALSE, include=TRUE, message=FALSE}
# Per-transect BB abundance
## full dataset
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  group_by(site, date, bb.sp) %>% # 
  summarize(bb.abund = n()) %>% # get per transect bb abundance
  ungroup() %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) %>% # complete implicit absences
  semi_join(net, by = c("site", "date")) %>% # drop site-dates that did not exist in the data set
  left_join(site_data) %>% # add site data
  mutate(year = factor(year(date)),
         yday = yday(date),
         bb.sp = factor(bb.sp),
         management = factor(management),
         site = factor(site),
         transect = factor(transect)) %>%
  group_by(year, bb.sp) %>%
  mutate(bb.annual.abund = sum(bb.abund))

@@

### Per-transect floral abund: by species
fl_abund_sp <- survey %>%
  group_by(site, date, plant.sp) %>% # group by plant.sp*transect
  summarize(fl.abund = sum(flower.cover)) %>% # get plant.sp*transect total flower cover; this step really shouldn't be required, but in a few cases, there is more than one abundance entry for the same species on a given transect; not sure how that happened, but it's in the original data that Fabrice sent me; alternatively, I could drop the dups or average them. We'll see what Kalli and Fabrice have to say.
  group_by(site, date) %>%
  mutate(prop.fl.abund = fl.abund/sum(fl.abund)) %>% # get the proportional flower cover for each species.
  ungroup() %>%
  #full_join(dplyr::select(net, site, date, plant.sp), by = c("site", "date", "plant.sp")) %>% # add the handful of species occurring in the visitation network but missing from the survey data
  complete(site, date, plant.sp, fill = list(fl.abund = 0, prop.fl.abund = 0)) %>%
  semi_join(survey, by = c("site", "date")) %>%  # only include site-dates that actually occurred
  unique() %>%
  mutate(year = factor(year(date)))
```

@@

#check.gamViz(weighted_fl_abundance.2010)
#weighted_fl_abundance.2010.sum <- summary(weighted_fl_abundance.2010)
#print(plot(weighted_fl_abundance.2010, allTerms = TRUE), pages = 4)
#weighted_fl_abundance.2010.ck <- check2D(weighted_fl_abundance.2010, x1 = "yday", x2 = "elev.mean")

@@

### Extract predictions
```{r message=FALSE}
# Function to build data set over which to generate predictions
newdata_build <- function(dat, y) {

@@

# Generate predictions
predict2010.FL <- predict_func(weighted_fl_abundance.2010, newdata2010.FL, 2010)
predict2010.FL.sqrt <- predict_func(weighted_fl_abundance.2010.sqrt, newdata2010.FL, 2010)
predict2010.FL.bin <- predict_func(weighted_fl_abundance.2010.bin, newdata2010.FL, 2010)

predict2011.FL <- predict_func(weighted_fl_abundance.2011, newdata2011.FL, 2011)
predict2011.FL.sqrt <- predict_func(weighted_fl_abundance.2011.sqrt, newdata2011.FL, 2011)
predict2011.FL.bin <- predict_func(weighted_fl_abundance.2011.bin, newdata2011.FL, 2011)

predict2012.FL <- predict_func(weighted_fl_abundance.2012, newdata2012.FL, 2012)
predict2012.FL.sqrt <- predict_func(weighted_fl_abundance.2012.sqrt, newdata2012.FL, 2012)
predict2012.FL.bin <- predict_func(weighted_fl_abundance.2012.bin, newdata2012.FL, 2012)

@@

predict_pool <- bind_rows(predict_pool.FL, predict_pool.BB)
predict_pool.sqrt <- bind_rows(predict_pool.FL.sqrt, predict_pool.BB)
predict_pool.bin <- bind_rows(predict_pool.FL.bin, predict_pool.BB)

# Subset predictions
predict_major <- filter(predict_pool, 
                           bb.sp %in% c("bss", "pasc", "prat", "soro", "wurf")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "bss" ~ "B. ss.",
    bb.sp == "pasc" ~ "B. pasc",
    bb.sp == "prat" ~ "B. prat",
    bb.sp == "soro" ~ "B. soro",
    bb.sp == "wurf" ~ "B. wurf"
  ))

predict_major.sqrt <- filter(predict_pool.sqrt, 
                           bb.sp %in% c("bss", "pasc", "prat", "soro", "wurf")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "bss" ~ "B. ss.",
    bb.sp == "pasc" ~ "B. pasc",
    bb.sp == "prat" ~ "B. prat",
    bb.sp == "soro" ~ "B. soro",
    bb.sp == "wurf" ~ "B. wurf"
  ))

predict_major.bin <- filter(predict_pool.bin, 
                           bb.sp %in% c("bss", "pasc", "prat", "soro", "wurf")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "bss" ~ "B. ss.",
    bb.sp == "pasc" ~ "B. pasc",
    bb.sp == "prat" ~ "B. prat",
    bb.sp == "soro" ~ "B. soro",
    bb.sp == "wurf" ~ "B. wurf"
  ))

predict_highelev <- filter(predict_pool, 
                              bb.sp %in% c("pyre", "mend", "mont")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "pyre" ~ "B. pyre",
    bb.sp == "mend" ~ "B. mend",
    bb.sp == "mont" ~ "B. mont"
  ))

predict_highelev.sqrt <- filter(predict_pool.sqrt, 
                                   bb.sp %in% c("pyre", "mend", "mont")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "pyre" ~ "B. pyre",
    bb.sp == "mend" ~ "B. mend",
    bb.sp == "mont" ~ "B. mont"
  ))

predict_highelev.bin <- filter(predict_pool.bin, 
                                  bb.sp %in% c("pyre", "mend", "mont")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "pyre" ~ "B. pyre",
    bb.sp == "mend" ~ "B. mend",
    bb.sp == "mont" ~ "B. mont"
  ))

predict_minor <- filter(predict_pool, 
                           bb.sp %in% c("gers", "hort", "hypn","jone", "lapi", "muci")) %>%
  mutate(bb.sp = case_when(
    bb.sp == "gers" ~ "B. gers",
    bb.sp == "hort" ~ "B. hort",
    bb.sp == "hypn" ~ "B. hypn",
    bb.sp == "jone" ~ "B. jone",
    bb.sp == "lapi" ~ "B. lapi",
    bb.sp == "muci" ~ "B. muci"
  ))

predict_minor.sqrt <- filter(predict_pool.sqrt, 
                                bb.sp %in% c("gers", "hort", "hypn","jone", "lapi", "muci")) %>% 
  mutate(bb.sp = case_when(
    bb.sp == "gers" ~ "B. gers",
    bb.sp == "hort" ~ "B. hort",
    bb.sp == "hypn" ~ "B. hypn",
    bb.sp == "jone" ~ "B. jone",
    bb.sp == "lapi" ~ "B. lapi",
    bb.sp == "muci" ~ "B. muci"
  ))

predict_minor.bin <- filter(predict_pool.bin, 
                               bb.sp %in% c("gers", "hort", "hypn","jone", "lapi", "muci")) %>% 
  mutate(bb.sp = case_when(
    bb.sp == "gers" ~ "B. gers",
    bb.sp == "hort" ~ "B. hort",
    bb.sp == "hypn" ~ "B. hypn",
    bb.sp == "jone" ~ "B. jone",
    bb.sp == "lapi" ~ "B. lapi",
    bb.sp == "muci" ~ "B. muci"
  ))
```

### Visualize
```{r}
#### Major
ggplot(predict_major, aes(date, elev.mean, fill = pred)) +
  geom_raster() +
  geom_contour(aes(z = pred), color = "black", size = 0.25) +
  scale_fill_viridis_c(name = "log(abund.)") +
  #facet_grid(bb.sp + subject ~ year) +
  facet_grid(year ~ bb.sp + subject) +
  xlab("Date") +
  ylab("Elevation (m)") +
  theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5))
ggsave("./output/GAM_major.png", height = 5, width = 11, units = "in")

>>>>

### FL and BB abundance / diversity
```{r echo=FALSE, include=TRUE, message=FALSE}
# Per-transect BB abundance
## full dataset
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  group_by(site, date, bb.sp) %>% # 
  summarize(bb.abund = n()) %>% # get per transect bb abundance
  ungroup() %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) %>% # complete implicit absences
  semi_join(net, by = c("site", "date")) %>% # drop site-dates that did not exist in the data set
  left_join(site_data) %>% # add site data
  mutate(year = factor(year(date)),
         yday = yday(date),
         bb.sp = factor(bb.sp),
         management = factor(management),
         site = factor(site),
         transect = factor(transect)) %>%
  group_by(year, bb.sp) %>%
  mutate(bb.annual.abund = sum(bb.abund),
         abund.class = case_when(
           bb.sp %in% c("bss", "pasc", "prat", "soro", "wurf") ~ "dominant",
           bb.sp %in% c("jone", "psit", "hort") ~ "common",
           bb.sp %in% c("pyre", "mend", "mont", "hypn", "lapi") ~ "high elev.",
           bb.sp %in% c("gers", "humi", "muci") ~ "minor"
         ),
         abund.class = factor(abund.class, levels = c("dominant", "common", "high elev.", "minor")),
         bb.sp.ord = factor(bb.sp, levels = c("bss", "pasc", "prat", "soro", "wurf", 
                                              "hort", "jone", "psit",
                                              "pyre", "mend", "mont",
                                              "gers", "humi", "muci", "hypn", "lapi")))

@@

### Per-transect floral abund: by species
fl_abund_sp <- survey %>%
  group_by(site, date, plant.sp) %>% # group by plant.sp*transect
  summarize(fl.abund = sum(flower.cover)) %>% # get plant.sp*transect total flower cover; this step really shouldn't be required, but in a few cases, there is more than one abundance entry for the same species on a given transect; not sure how that happened, but it's in the original data that Fabrice sent me; alternatively, I could drop the dups or average them. We'll see what Kalli and Fabrice have to say.
  group_by(site, date) %>%
  mutate(prop.fl.abund = fl.abund/sum(fl.abund)) %>% # get the proportional flower cover for each species.
  ungroup() %>%
  #full_join(dplyr::select(net, site, date, plant.sp), by = c("site", "date", "plant.sp")) %>% # add the handful of species occurring in the visitation network but missing from the survey data
  complete(site, date, plant.sp, fill = list(fl.abund = 0, prop.fl.abund = 0)) %>%
  semi_join(survey, by = c("site", "date")) %>%  # only include site-dates that actually occurred
  unique() %>%
  mutate(year = factor(year(date)))

### Visualize raw abundance/occurrence
ggplot(filter(bb_abund_sp, bb.abund > 0), aes(yday, elev.mean, color = abund.class)) +
  geom_point(aes(alpha = bb.abund)) +
  #geom_point() +
  #facet_grid(year~ bb.sp) 
  facet_wrap(~bb.sp.ord)

ggplot(bb_abund_sp, aes(bb.sp, log(bb.abund + 1))) +
  geom_boxplot()

ggplot(bb_abund_sp, aes(log(bb.abund + 1))) +
  geom_histogram() +
  facet_wrap(~bb.sp)
```

@@

#check.gamViz(weighted_fl_abundance.2010)
weighted_fl_abundance.2010.sum <- summary(weighted_fl_abundance.2010)
saveRDS(weighted_fl_abundance.2010.sum, file = "./output/weighted_fl_abundance_2010.rds")
#print(plot(weighted_fl_abundance.2010, allTerms = TRUE), pages = 4)
#weighted_fl_abundance.2010.ck <- check2D(weighted_fl_abundance.2010, x1 = "yday", x2 = "elev.mean")

@@

### Make a data set that sets the boundaries for plotting predictions extrapolated (or interpolated) beyond samples
```{r}
survey_samples <- survey %>%
  ungroup() %>%
  left_join(site_data) %>%
  mutate(yday = yday(date),
         year = year(date)) %>%
  select(year, yday, elev.mean) %>%
  unique() %>%
  mutate(sample = factor(c(1:length(year)))) %>%
  group_by(year, sample) %>%
  arrange(yday)
%>%
  nest()

dat <- survey_samples$data[[1]]

func <- function(x, days, meters) {
  for(i in 1:dim(x)[1]) {

    yday = x[i, 2]
    print(yday)
   # x = seq(x$)
  }
}

func(survey_samples),

dat <- survey_samples$data[[1]]
yday <- dat$yday # set reference yday
ydays <- seq((yday - 7), (yday + 7), 0.5) # add all days within range
elev <- dat$elev.mean # set reference elev
elevs <- seq((elev - 100), (elev + 100), 1) # add all elevs within range
out.temp <- expand.grid(ydays, elevs) # expand grid to all possible combos

bb_samples <- net %>%
  ungroup() %>%
  left_join(site_data) %>%
  mutate(yday = yday(date),
         year = year(date)) %>%
  select(year, yday, elev.mean) %>%
  unique() %>%
  mutate(sample = factor(c(1:length(year)))) %>%
  group_by(sample, year) %>%
  nest()

### Fill func
#### Creates a set of yday*elevation combinations that can be used to filter surface plots to exclude predictions too far from the data
#### This is very hard-wired right now, and it would break if any of the column names changed.
plotting_boundaries <- function(x,
                                year_sub, # a sloppy way to let me filter by year
                                days, # how many days of outside our samples will we plot?
                                meters) { # how many meters of outside our samples will we plot?

  #out <- tibble() # create empty tibble to fill
  out <- c()

  data <- x %>%
    filter(year == year_sub)

  for(i in 1:dim(data)[1]) {

    dat <- x$data[[i]] # this pull out the ith data frame of the data column in the nested data frame

    yday <- dat$yday # set reference yday
    # ydays <- seq((yday - days), (yday + days), 0.5) # add all days within range
    # 
    # elev <- dat$elev.mean # set reference elev
    # elevs <- seq((elev - meters), (elev + meters), 1) # add all elevs within range
    # 
    # out.temp <- expand.grid(ydays, elevs) # expand grid to all possible combos
    # out <- bind_rows(out, out.temp)
    out[i] <- yday
  }
  # out <- unique(out) %>% # drop duplicates created by overlapping ranges
  #   select(yday = 1, elev.mean = 2) # name columns
  out
}

FL_plotting_2010 <- plotting_boundaries(survey_samples, 2010, 7, 100)
FL_plotting_2011 <- plotting_boundaries(survey_samples, 2011, 7, 100)
FL_plotting_2012 <- plotting_boundaries(survey_samples, 2012, 7, 100)
View(unique(FL_plotting_2012$yday))

BB_plotting_2010 <- plotting_boundaries(bb_samples, 2010, 7, 100)
BB_plotting_2011 <- plotting_boundaries(bb_samples, 2011, 7, 100)
BB_plotting_2012 <- plotting_boundaries(bb_samples, 2012, 7, 100)
```

### Extract predictions
```{r message=FALSE}
### Species names and such for plotting
species_names <- cbind(c("bss", "pasc", "prat", "soro", "wurf", 
                         "hort", "pyre", "mend", "mont", "gers", 
                         "hypn","jone", "lapi", "muci", "psit", "himi"),
                       c("B. sensu-strictu", "B. pascuorum", "B. pratorum", 
                         "B. soroensis", "B. wurflenii", "B. hortorum",
                         "B. pyrenaeus", "B. mendax", "B. monticola",
                         "B. gerstaeckeri", "B. hypnorum", "B. jonellus", 
                         "B. lapidarius", "B. mucidus", "B. psithyrus", "B. humilis"),
                       c("B. ss.", "B. pasc.", "B. prat.", 
                         "B. soro.", "B. wurf.", "B. hort.",
                         "B. pyre.", "B. mend.", "B. mont.",
                         "B. gers.", "B. hypn.", "B. jone.", 
                         "B. lapi.", "B. muci.", "B. psit.", "B. humi"),
                       c("major", "major", "major", "major", "major",
                         "major", "high elev.", "high elev.", "high elev.",
                         "minor", "minor", "minor", "minor", "minor", "minor", "minor")
                       ) %>%
  as.data.frame() %>%
  as_tibble() %>%
  dplyr::select(bb.sp = 1, species.name = 2, species.abb = 3, abund.class = 4)
# Function to build data set over which to generate predictions
newdata_build <- function(dat, y) {

@@

# Generate predictions
predict2010.FL <- predict_func(weighted_fl_abundance.2010, newdata2010.FL, 2010) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2010, by =  c("yday", "elev.mean"))
predict2010.FL.sqrt <- predict_func(weighted_fl_abundance.2010.sqrt, newdata2010.FL, 2010) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2010, by =  c("yday", "elev.mean"))
predict2010.FL.bin <- predict_func(weighted_fl_abundance.2010.bin, newdata2010.FL, 2010) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2010, by =  c("yday", "elev.mean"))

predict2011.FL <- predict_func(weighted_fl_abundance.2011, newdata2011.FL, 2011) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2011, by =  c("yday", "elev.mean"))
predict2011.FL.sqrt <- predict_func(weighted_fl_abundance.2011.sqrt, newdata2011.FL, 2011) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2011, by =  c("yday", "elev.mean"))
predict2011.FL.bin <- predict_func(weighted_fl_abundance.2011.bin, newdata2011.FL, 2011) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2011, by =  c("yday", "elev.mean"))

predict2012.FL <- predict_func(weighted_fl_abundance.2012, newdata2012.FL, 2012)
predict2012.FL2 <- predict2012.FL %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2012, by =  c("yday", "elev.mean"))

ggplot(predict2012.FL2, aes(yday, elev.mean, fill = pred)) +
  geom_raster() +
  geom_contour(aes(z = pred), color = "black", size = 0.25) +
  scale_fill_viridis_c(name = "log(abund.)") +
  facet_wrap(~ bb.sp)

predict2012.FL.sqrt <- predict_func(weighted_fl_abundance.2012.sqrt, newdata2012.FL, 2012) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2012, by =  c("yday", "elev.mean"))
predict2012.FL.bin <- predict_func(weighted_fl_abundance.2012.bin, newdata2012.FL, 2012) %>%
  mutate(yday = ceiling(yday*2)/2) %>%
  semi_join(FL_plotting_2012, by =  c("yday", "elev.mean"))

@@

predict_pool <- bind_rows(predict_pool.FL, predict_pool.BB) %>%
  left_join(species_names)
predict_pool.sqrt <- bind_rows(predict_pool.FL.sqrt, predict_pool.BB) %>%
  left_join(species_names)
predict_pool.bin <- bind_rows(predict_pool.FL.bin, predict_pool.BB) %>%
  left_join(species_names)
```

### Visualize
```{r}
#### Major
ggplot(filter(predict_pool, abund.class == "major"), aes(date, elev.mean, fill = pred)) +
  geom_raster() +
  geom_contour(aes(z = pred), color = "black", size = 0.25) +
  scale_fill_viridis_c(name = "log(abund.)") +
  #facet_grid(bb.sp + subject ~ year) +
  facet_grid(year ~ bb.sp + subject) +
  xlab("Date") +
  ylab("Elevation (m)") +
  theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5))
ggsave("./output/GAM_major.png", height = 3, width = 6.5, units = "in")
